Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-10 writeup/pnas-word-order-si.tex

@@

>>>>

\paragraph{Evaluating Accuracy of Formalization}
An anonymous reviewer notes that the mapping between Dryer's relations and UD is not perfect, since some of the UD relations subsume other relations.
Here we show that this is not a problem, since the ordering of the various relations subsumed under the UD label typologically strongly agree.

\begin{enumerate}
	\item Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}} captures correlations with inflected tense, aspect, and negation auxiliaries as stated by \citet{dryer1992greenbergian}; however, \emph{aux} aso encompasses other types of auxiliaires, such as modals.
We note that other authors, including \citet{greenberg1963universals}, have stated the correlation for all inflected auxiliaries; for further references, we refer to \citet[Number 501]{plank2000universals}.

We used the UD treebanks to confirm that different auxiliaires tend to pattern together, and that the most frequent order of the \emph{aux} relation coincides with that of inflected tense-aspect or negation auxiliaries.

We collected, for each UD language, all dependents of the \emph{aux} dependency, occurring at least 10 times, and compared their dominant orders, which we operationalized as their more common order in the treebank (auxiliary-head or head-auxiliary).
The dependency occurs in all but two very small treebanks (Telugu and Irish).
In 43 languages, all extracted auxiliaries had the same dominant order, with the possible exception of uninflected particles labeled \emph{aux} (Croatian, German, Polish, Ukrainian).
In three languages (Ancient Greek, Russian), there were other auxiliaries with different dominant order, but these were modal or passive auxiliaries.
Finally, in three languages (Afrikaans, Old Church Slavonic, and Persian), not all tense-aspect auxiliaries showed the same dominant order as the \emph{aux} dependency overall.
For instance, in Persian, the perfect auxiliary \emph{budan} follows the main verb, whereas the future auxiliary \emph{xaastan xaah-} precedes it~\citep[pp. 117, 121]{mace2015persian}.
%		(1) In Afrikaans, the past auxiliary \emph{het} shows dominant aux-head order, whereas the overall most common order of auxiliaries is head-aux (e.g. future auxiliary \emph{sal}).
%		Finally, in Old Church Slavonic, most aux dependencies have the reflexive \emph{sebe} which tends to precede the head, whereas inflected verbal auxiliaries (primarily, forms of \emph{byti}) have opposite order, leading to a seeming violation of the correlation  \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}.

Taken together, this shows that the dominant order of the \emph{aux} relation strongly coincides with that of inflected tense-aspect auxiliaries, except for a small number of languages where different tense-aspect auxiliaries show different orders.

\item Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}} is formalized using \emph{nmod} which covers not only genitives, but also all other noun-modifying NPs and PPs.
	The evaluation of extracted grammars against WALS (Table~\ref{tab:grammars-wals}) shows that, among the 37 languages where WALS has an entry, the dominant direction of \emph{nmod} agrees with that of genitives, with two exceptions (Danish and Swedish).

TODO figure out why

\item Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}} is formalized using \emph{acl}, which covers not just relative clauses, but also other adnominal clauses.
	In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{acl} agrees with the WALS entry for relative clauses in all but three languages (Estonian, Finnish, Tamil) out of the 36 languages for which WALS has an entry.
		Also, UD provides a specific \emph{acl:relcl} sub-label for relative clauses in 21 of the languages.
		In all but three languages, the dominant order is the same for the general \emph{acl} label as for the specific \emph{acl:relcl} one (exceptions: Estonian, Finnish, Hindi).

TODO figure out why

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}} is formalized using \emph{nmod}, which covers not only PPs and standards of comparison, but also adjunct NPs.
		In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{nmod} agrees with that annotated for obliques in all 18 languages for which WALS has an entry.

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}} is formalized using \emph{xcomp}, which covers other control verbs, not just verbs of volition (`want').

		We used the UD treebanks to investigate whether there are differences in the ordering of `want' and other verbs using the \emph{xcomp} dependency.

		The dependency is annotated in all but two languages (Japanese and Turkish).

		For each language, we extracted all lemmas of words heading an \emph{xcomp} dependency, occurring at least 2 times.
In 39 languages, all extracted words had the same dominant order.
Additionally, in four Germanic languages (Afrikaans, Danish, Dutch, and German), the verb of volition (Afrikaans \emph{wil}, Danish \emph{ville}, Dutch \emph{willen}, German \emph{wollen}) is mostly annotated with the \emph{aux} relation due to UD annotation guidelines, but in all languages, its dominant order (verb of volition before its complement) agrees with the dominant order of the \emph{xcomp} dependency (head-initial).
		In three historical languages (Ancient Greek, Latin, and Old Church Slavonic), verbs of volition agree with the dominant order of \emph{xcomp}, while several other verbs that do not indicate volition show opposite dominant order.
Finally, in Gothic, the verb of volition (\emph{wiljan}) has opposite order, resulting in an apparent violation of Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}.

		Taken together, the order of the xcomp dependency tends to agree with that of most other \emph{xcomp} dependencies, with the sole exception of Gothic.
\end{enumerate}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-raw.pdf}
	\caption[Predictability and Parseability]{Raw numerical values estimated for Predictability (negative surprisal), and negative syntactic ambiguity $-H[T|U]$, before $z$-scoring. For more meaningful comparison, both quantities are normalized by the number of words in the corpus, i.e., we plot per-word negative surprisal and per-word difficulty in recovering syntactic structures.  Note that the negative syntactic ambiguity $-H[T|U]$ equals parseability $I[T,U] = H[T] - H[T|U]$ up to a per-language constant $H[T]$, which we do not attempt to estimate.}\label{fig:pareto-per-lang-raw}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized.pdf}
	\caption[Predictability and Parseability]{Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}
\end{figure}
